---
title: "urban_gauge2"
author: "John Gardner"
date: "July 29, 2020"
output: html_document
---

```{r setup, include=FALSE}
# LOAD packages
library(stringr)
library(tidyverse)
library(feather)
library(sf)
library(rgdal)
library(maps)
library(magrittr)
library(purrr)
library(data.table)
library(tmap)
library(readr)
library(lubridate)
#library(raster)
library(mapview)
library(ggthemes)
library(vroom) 
library(leaflet)
library(tidyr)
library(leafpop)
library(USAboundaries)
library(tigris)
library(tidycensus)
library(knitr)
library(dataRetrieval)
library(plotly)

knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r census_API, include=F, eval=T, echo=F}

### get census API key at  http://api.census.gov/data/key_signup.html

### Run this only first time
#census_api_key("739f668cac5542524a7b3a5e3977032eb3319330", overwrite = T, install =T)

###  First time, reload your environment so you can use the key without restarting R.
#readRenviron("~/.Renviron")

###  You can check it with:And use this to re-load key in future sessions
#Sys.getenv("CENSUS_API_KEY")


```


```{r census_data, echo=F, include=F}
#### get census data

### find which variable is total population
#v17 <- load_variables(2017, "acs1", cache = TRUE)
#pop_v17 <- v17 %>% 
#  filter(grepl("POP", concept)) %>% 
#  as.data.frame()

### download total population estimates for urban boundaries >65000 by year
#years <- lst( 2005, 2006, 2007,2008,2009,2010,2011,2012,2013,2014,2015,2016,2017,2018)

# pop_urban <- purrr::map_dfr(
#   years,
#   ~ get_acs(
#     geography = "urban area",
#     variable = "B01003_001",
#     year = .x,
#     survey = "acs1",
#     geometry = FALSE
#     ),
#   .id="year"
# )

#download urban area simple geometry
urban <- urban_areas(cb=T, class="sf")

pop_urban<-read_csv("out/urban_areas_pop_2005_2018.csv")

# https://www.census.gov/programs-surveys/geography/guidance/geo-areas/urban-rural/2010-urban-rural.html
```

# To get a sense of the spatial distribution the top 100 largest urban areas (mean 2005-2018 pop), other urban areas (defined by census as pop > 65,000), and urban clusters (any incorporated town, village, or city with pop >2,500). 

```{r map, echo=F}

# calcualte annual rate of change of populations and median over time
pop_urban_sum <- pop_urban %>%
  mutate(year = as.numeric(year)) %>%
  group_by(GEOID) %>%
  arrange(year) %>%
  mutate(mean_pop= mean(estimate, na.rm=T),
         lag_pop = lag(estimate),
         pct_change = ((estimate-lag_pop)/lag_pop) *100) %>%
  mutate(mean_pct_change = mean(pct_change, na.rm=T),
         median_pct_change = median(pct_change, na.rm = T))  %>%
  ungroup() %>%
  arrange(GEOID, year)

# filter to top 100 largest cities (mean pop 2005-2018)
pop_plot <- pop_urban_sum %>%
  group_by(GEOID) %>%
  filter(year == max(year, na.rm = T)) %>%
  ungroup() %>%
  arrange(desc(mean_pop)) %>%
  slice(1:100) %>%
  mutate(
    population = case_when(
      median_pct_change < 0 ~ "Decreasing",
      median_pct_change > 0 ~ "Increasing",
      median_pct_change == 0 ~ "No Change"))


# plot all urban clusters, areas, and the largest 100
mapview(urban %>% filter(!GEOID10 %in% pop_urban$GEOID), col.regions="yellow", alpha=0.5, layer.name="Urban Clusters (>2500 ppl)") +
mapview(urban %>%
            filter(GEOID10 %in% unique(pop_urban$GEOID)), layer.name="Urban Areas (>65,000 ppl)", col.regions="orange", alpha=0.5) +
mapview(urban %>%
          inner_join(pop_plot, by=c("GEOID10"="GEOID")),
                     col.regions="red", layer.name="Top 100 largest urban areas", alpha=0.5) 


```
# First lets look at how 8586 active (and recently active) USGS gages overlap with all 3601 urban clusters/areas. 

# How many gages are WITHIN these "urban" boundaries (pop >2500) ?
# 24% of gages are within urban areas, and 76% percent are in rural areas

# To compare this with urban vs. rural population and area
# 80.7 % of US population lives in urban areas, but urban area accounts for 2.9% of the total land area

# How many urban areas (pop > 2500) have at least 1 gage?
# 81% of urban areas have NO gages, 19% have at least 1 gage. Ones without gages are generally small towns as we will see later.
# This seems like problem for water resource management and flood risk for small rural towns that may not have a gage nearby.

```{r intersect_UC, echo=F}

## load usgs gauges (8587
gauge <- read_csv("in/All-USGS-v1.csv") %>%
  filter(!is.na(latitude) & !is.na(longitude)) %>%
  st_as_sf(coords= c("longitude","latitude"), crs=4269)
 
### FIND gauges IN all urban boundaries. 
UC_join <- gauge %>%
  st_transform(2163) %>%
  st_join(urban %>%
            st_transform(2163),
          join=st_intersects)

### interactive map of all gauges colored by city they are in
mapview(UC_join, zcol="GEOID10", legend=F, layer.name= "Gages") +
  mapview(urban, zcol="GEOID10", legend=F, layer.name="All Urban areas/clusters")

# estimate percent of all gauges in all urban areas/clusters
knitr::kable(UC_join %>%
  st_set_geometry(NULL) %>%
  mutate(urban_gauge = ifelse(is.na(GEOID10), "No","Yes")) %>%
  group_by(urban_gauge) %>%
  summarise(Count = n()) %>%
  mutate(Percent = Count/sum(Count)*100),
  digits=2, align="c", caption= "# gauges WITHIN urban areas/clusters >2,500 people")

# estimate how many urban areas/clusters have gauges or not
UC_no_gage <- urban %>%
  st_set_geometry(NULL) %>%
  left_join(UC_join %>%
              st_set_geometry(NULL) %>%
              select(GEOID10) %>%
              distinct(GEOID10) %>%
              mutate(has_gauge ="YES"),
            by="GEOID10") %>%
  mutate(has_gauge = ifelse(is.na(has_gauge), "NO", has_gauge))

knitr::kable(UC_no_gage %>%
  group_by(has_gauge) %>%
  summarise(Count=n()) %>%
  mutate(Percent = Count/sum(Count)*100),
  digits=2, align="c", caption="# urban areas/clusters with gauges")

#total land area of USA, alaska, HI, 
#8080464.3 + 1717856 + 28311

#(sum(urban$ALAND10)/1000000+ sum(urban$AWATER10)/1000000) / (8080464.3 + 1717856 + 28311)

```

# Does the number of gages scale with land area of cities?Probably Yes

# But what about all those urban areas with NO gages, are they smaller on average? Yes, but big cities can have 0 gages, and small towns can have 2-4 gages

```{r area_scaling, echo=F}

#### of urban areas with gages, is the # of gages relates to land area
p3 <- UC_join %>%
  st_set_geometry(NULL) %>%
  filter(!is.na(GEOID10)) %>%
 # mutate(urban_gage = ifelse(is.na(GEOID10), "No","Yes")) %>%
  group_by(GEOID10) %>%
  summarise(Count = n(), ALAND10=ALAND10[1], site_years=sum(rec_length)) %>%
  mutate(Percent = Count/sum(Count)*100) %>%
  inner_join(urban %>%
                st_set_geometry(NULL) %>%
                select(NAME10, GEOID10), by="GEOID10")%>%
  arrange(desc(site_years)) %>%

ggplot() +
  geom_point(aes(x=ALAND10/1000000, y=Count, label=NAME10, color=log10(site_years))) +
  scale_color_viridis_c()+
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  xlab("Size of urban area (km2)")+
  ylab("Number of gages")

ggplotly(p3)

###
urban %>%
  st_set_geometry(NULL) %>%
  left_join(UC_join %>%
              st_set_geometry(NULL) %>%
              dplyr::select(sitename, drainage.area,rec_length, GEOID10) %>%
              dplyr::filter(!is.na(GEOID10)) %>%
              mutate(has_gauge ="YES"),
            by="GEOID10") %>%
  mutate(has_gauge = ifelse(is.na(has_gauge), "NO", has_gauge)) %>%
  group_by(GEOID10) %>%
  summarise(gage_count = sum(has_gauge=="YES"), 
            ALAND10=ALAND10[1], site_years=sum(rec_length)) %>%
  mutate(num_gages = case_when(
    is.na(gage_count)|gage_count==0 ~ "0",
    gage_count ==1 ~ "1",
    gage_count >1 & gage_count <= 4 ~ "2-4",
    gage_count >4 & gage_count <= 10 ~ "5-10",
    gage_count >10  ~ "Greater than 10"
  )) %>%

  
ggplot()+
  geom_density(aes(ALAND10/1000000, fill=num_gages), alpha=0.5)+
  scale_x_log10() +
 # scale_y_log10()+
  scale_fill_viridis_d()+
  theme_bw()+
  xlab("Size of urban area (km2)")

```

# Next, lets focus on the 447 locations the census defines as "urban areas" i.e. with pop >65,000 (because we have easily accesible population data over time for these areas).

# How many of the 8586 gages are WITHIN cities (pop >65,000) ?
# 19% of gages are within these cities, and 81% are outside of these cities


# How many cities (pop > 65,000) have at least 1 gage?
# 36% of cities do NOT have any USGS gages in the boundaries. 64% of citees have at least one gage (that was active in last ~7 years).

```{r intersect_UA, echo=F}

### FIND gauges IN urban boundaries. 
# filter urban areas to ones with pop data first.
UA_join <- gauge %>%
  st_transform(2163) %>%
  st_join(urban %>%
            filter(GEOID10 %in% unique(pop_urban$GEOID)) %>%
            st_transform(2163),
          join=st_intersects)

### 447 areas
urban %>%
  st_set_geometry(NULL) %>%
  filter(GEOID10 %in% unique(pop_urban$GEOID)) %>%
  summarise(n=length(unique(GEOID10)))

### interactive map of all gauges colored by city they are in
mapview(UA_join, zcol="GEOID10", legend=F, layer.name= "Gauges") +
  mapview(urban %>%
            filter(GEOID10 %in% unique(pop_urban$GEOID)), zcol="GEOID10", legend=F, layer.name="Urban Boundaries")

# estimate percent of all gauges in urban areas
knitr::kable(UA_join %>%
  st_set_geometry(NULL) %>%
  mutate(urban_gauge = ifelse(is.na(GEOID10), "No","Yes")) %>%
  group_by(urban_gauge) %>%
  summarise(Count = n()) %>%
  mutate(Percent = Count/sum(Count)*100),
  digits=2, align="c", caption= "# gauges IN cities >65,000 people" )

# estimate how many cities have gauges or not
UA_no_gage <- urban %>%
  filter(GEOID10 %in% unique(pop_urban$GEOID)) %>%
  st_set_geometry(NULL) %>%
  left_join(UA_join %>%
              st_set_geometry(NULL) %>%
              select(GEOID10) %>%
              distinct(GEOID10) %>%
              mutate(has_gage ="YES"),
            by="GEOID10") %>%
  mutate(has_gage = ifelse(is.na(has_gage), "NO", has_gage))

knitr::kable(UA_no_gage %>%
  group_by(has_gage) %>%
  summarise(Count=n()) %>%
  mutate(Percent = Count/sum(Count)*100),
  digits=2, align="c", caption="# cities with gauges")

```

# We showed there may be scaling between land area and # of gages.. How about population and # of gages (for cities > 65000)? Yes, more populous cities have more gages.
# Hover over the scattplot to know which cities plot where. Looks like three outliers with high population but few gages (Miami, Phoenix, LA). These are weird places (have fewer rivers and probably water districts/state that collect that data)

```{r pop_scaling, echo=F}
# pop_mean <- pop_urban %>%
#   group_by(GEOID) %>%
#   summarise(mean_pop= mean(estimate, na.rm=T)) %>%
#   ungroup()

p1 <-UA_join %>%
  st_set_geometry(NULL) %>%
  filter(!is.na(GEOID10)) %>%
 # mutate(urban_gage = ifelse(is.na(GEOID10), "No","Yes")) %>%
  group_by(GEOID10) %>%
  summarise(Count = n(), 
            ALAND10=ALAND10[1], 
            site_years=sum(rec_length)) %>%
 inner_join(pop_mean, by=c('GEOID10'='GEOID')) %>%
  inner_join( urban %>%
                st_set_geometry(NULL) %>%
                select(NAME10, GEOID10), by="GEOID10")%>%

ggplot(aes(x=mean_pop, y=Count, color=log10(ALAND10/1000000),
             label=NAME10)) +
  geom_point() +
  scale_color_viridis_c()+
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  xlab("Mean population")+
  ylab("Number of gages")+
  labs(color="Log10(Area)") 
 # geom_text(aes(label=NAME10),hjust=0, vjust=0, color="grey", alpha=0.5)

ggplotly(p1)

###
urban %>%
  filter(GEOID10 %in% unique(pop_urban$GEOID)) %>%
  st_set_geometry(NULL) %>%
  left_join(UA_join %>%
              st_set_geometry(NULL) %>%
              dplyr::select(sitename, drainage.area,rec_length, GEOID10) %>%
              dplyr::filter(!is.na(GEOID10)) %>%
              mutate(has_gauge ="YES"),
            by="GEOID10") %>%
  mutate(has_gauge = ifelse(is.na(has_gauge), "NO", has_gauge)) %>%
  group_by(GEOID10) %>%
  summarise(gage_count = sum(has_gauge=="YES"), 
            ALAND10=ALAND10[1], site_years=sum(rec_length)) %>%
  left_join(pop_mean, by=c('GEOID10'='GEOID')) %>%
  mutate(num_gages = case_when(
    is.na(gage_count)|gage_count==0 ~ "0",
    gage_count ==1 ~ "1",
    gage_count >1 & gage_count <= 4 ~ "2-4",
    gage_count >4 & gage_count <= 10 ~ "5-10",
    gage_count >10  ~ "Greater 10"
  )) %>%

  
ggplot()+
  geom_density(aes(mean_pop, fill=num_gages), alpha=0.5)+
  scale_x_log10() +
 # scale_y_log10()+
  scale_fill_viridis_d()+
  theme_bw()+
  xlab("Mean population")
```


# Lets look at cities (pop >65,000) at have increasing or decreasing pop (according to the median rate of year to year population change).

# %20 of cities have ~decreasing pop, 80% of cities have ~increasing pop. The relationshop between number of gages and city population is the same in cities with increasing and decreasing populations.
```{r city_pop, echo=F}
 
# make figs map pop increasing or decreasing

pop_urban_sum <- pop_urban %>%
  mutate(year = as.numeric(year)) %>%
  group_by(GEOID) %>%
  arrange(year) %>%
  mutate(mean_pop= mean(estimate, na.rm=T),
         lag_pop = lag(estimate),
         pct_change = ((estimate-lag_pop)/lag_pop) *100) %>%
  mutate(mean_pct_change = mean(pct_change, na.rm=T),
         median_pct_change = median(pct_change, na.rm = T))  %>%
  ungroup() %>%
  arrange(GEOID, year)

pop_plot <- pop_urban_sum %>%
  group_by(GEOID) %>%
  filter(year == max(year, na.rm = T)) %>%
  ungroup() %>%
  arrange(desc(mean_pop)) %>%
  #slice(1:100) %>%
  mutate(
    population = case_when(
      median_pct_change < 0 ~ "Decreasing",
      median_pct_change > 0 ~ "Increasing",
      median_pct_change == 0 ~ "No Change"))

## 20/80% decreasing/increasing pop
# pop_plot %>%
#   group_by(population) %>%
#   summarise(n=n())

# make map of increasiong/decreasing city pop
mapview(urban %>%
          inner_join(pop_plot, by=c("GEOID10"="GEOID")),
                     zcol="population", layer.name="Median % pop change", alpha=0.5)  
  

###
p2 <- UA_join %>%
  st_set_geometry(NULL) %>%
  filter(!is.na(GEOID10)) %>%
 # mutate(urban_gage = ifelse(is.na(GEOID10), "No","Yes")) %>%
  group_by(GEOID10) %>%
  summarise(Count = n(), 
            ALAND10=ALAND10[1], 
            site_years=sum(rec_length)) %>%
 inner_join(pop_plot, by=c('GEOID10'='GEOID')) %>%
 inner_join(urban %>%
                st_set_geometry(NULL) %>%
                select(NAME10, GEOID10), by="GEOID10")%>%
  filter(!is.na(population)) %>%

ggplot(aes(x=mean_pop, y=Count, color=population,
             label=NAME10)) +
  geom_point( alpha=0.4) +
  scale_color_viridis_d()+
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  xlab("Mean population")+
  ylab("Number of gages")

ggplotly(p2)


#  mapview(gauge_urban_join %>%
#            filter(GEOID10 %in% pop_plot$GEOID), col="black", #col.regions="black", alpha=0.5, cex=0.7, layer.name="Urban Gauges")

```

# Now lets focus on the top 100 largest cities.

# 93 of largest cities have at least 1 gage, 7 have no gages.
```{r top_100, echo=F}

pop_100 <- pop_urban_sum %>%
  group_by(GEOID) %>%
  filter(year == max(year, na.rm = T)) %>%
  ungroup() %>%
  arrange(desc(mean_pop)) %>%
  slice(1:100) %>%
  mutate(
    population = case_when(
      median_pct_change < 0 ~ "Decreasing",
      median_pct_change > 0 ~ "Increasing",
      median_pct_change == 0 ~ "No Change"))

### FIND gauges IN urban boundaries. 
# filter urban areas to ones with pop data first.
top100_join <- gauge %>%
  st_transform(2163) %>%
  st_join(urban %>%
            filter(GEOID10 %in% unique(pop_100$GEOID)) %>%
            st_transform(2163),
          join=st_intersects)

### 96 areas
# urban %>%
#   st_set_geometry(NULL) %>%
#   filter(GEOID10 %in% pop_100$GEOID) %>%
#   summarise(n=length(unique(GEOID10)))

### interactive map of all gauges colored by city they are in
mapview(top100_join, zcol="GEOID10", legend=F, layer.name= "Gages") +
  mapview(urban %>%
            filter(GEOID10 %in% unique(pop_100$GEOID)), zcol="GEOID10", legend=F, layer.name="Top 100 largest cities")


# estimate percent of all gauges in urban areas
knitr::kable(top100_join %>%
  st_set_geometry(NULL) %>%
  mutate(urban_gauge = ifelse(is.na(GEOID10), "No","Yes")) %>%
  group_by(urban_gauge) %>%
  summarise(Count = n()) %>%
  mutate(Percent = Count/sum(Count)*100),
  digits=2, align="c", caption= "# gauges IN top 100 largest cities" )


# estimate how many cities have gauges or not
top_no_gage <- urban %>%
  filter(GEOID10 %in% unique(pop_100$GEOID)) %>%
  st_set_geometry(NULL) %>%
  left_join(top100_join %>%
              st_set_geometry(NULL) %>%
              select(GEOID10) %>%
              distinct(GEOID10) %>%
              mutate(has_gage ="YES"),
            by="GEOID10") %>%
  mutate(has_gage = ifelse(is.na(has_gage), "NO", has_gage))

knitr::kable(top_no_gage %>%
  group_by(has_gage) %>%
  summarise(Count=n()) %>%
  mutate(Percent = Count/sum(Count)*100),
  digits=2, align="c", caption="# of top 100 cities with gauges")

knitr::kable(top_no_gage %>%
 filter(has_gage=="NO") %>%
 dplyr::select(NAME10, has_gage),
  digits=2, align="c", caption="Big cities with NO gages")



```






```{r buffer, include=F, echo=F}

### FIND gauges IN 1 km BUFFER boundaries. 

# # filter urban areas to ones with pop data first.
# gauge_urban_join <- gauge %>%
#   st_transform(2163) %>%
#   st_join(urban %>%
#             filter(GEOID10 %in% unique(pop_urban$GEOID)) %>%
#             st_transform(2163),
#           join=st_intersects)
# 
# ### interactive map of all gauges colored by city they are in
# mapview(gauge_urban_join, zcol="GEOID10", legend=F) +
#   mapview(urban %>%
#             filter(GEOID10 %in% unique(pop_urban$GEOID)), zcol="GEOID10", legend=F)
# 
# # estimate percent of all gauges in urban areas
# knitr::kable(gauge_urban_join %>%
#   st_set_geometry(NULL) %>%
#   mutate(urban_gauge = ifelse(is.na(GEOID10), "No","Yes")) %>%
#   group_by(urban_gauge) %>%
#   summarise(Count = n()) %>%
#   mutate(Percent = Count/sum(Count)*100),
#   digits=2, align="c", caption= "# gauges IN cities >65,000 people" )
# 
# 
# cities_wo_gauge <- urban %>%
#   filter(GEOID10 %in% unique(pop_urban$GEOID)) %>%
#   st_set_geometry(NULL) %>%
#   left_join(gauge_urban_join %>%
#               st_set_geometry(NULL) %>%
#               select(GEOID10) %>%
#               distinct(GEOID10) %>%
#               mutate(has_gauge ="YES"),
#             by="GEOID10") %>%
#   mutate(has_gauge = ifelse(is.na(has_gauge), "NO", has_gauge))
# 
# 
# knitr::kable(cities_wo_gauge %>%
#   group_by(has_gauge) %>%
#   summarise(Count=n()) %>%
#   mutate(Percent = Count/sum(Count)*100),
#   digits=2, align="c", caption="# cities with gauges")
# 
# 
# write_csv(pop_urban, "out/urban_areas_pop_2005_2018.csv")
# 
# st_write(urban, "out/urban_areas_boundaries.shp")
# 
# write_csv(gauge_urban_join %>%
#             filter(!is.na(GEOID10)), "out/gauges_IN_cities.csv")



```




```{r save, include=F, echo=F}


#write_csv(pop_urban, "out/urban_areas_pop_2005_2018.csv")

#st_write(urban, "out/urban_areas_boundaries.shp")

#write_csv(gauge_urban_join %>%
#            filter(!is.na(GEOID10)), "out/gauges_IN_cities.csv")


```

